Supplementary material for "Hack Weeks as a model for Data Science Education and Collaboration"

### 2018-06-04 conclusions.tex

The fast-paced changes of the computational and methodological landscape require traditional fields of science to rapidly adapt to new data analysis challenges.
To address these challenges, new types of workshops, including unconferences, hackathons and bootcamps, have been developed in recent years in various scientific disciplines to exist alongside with and support the existing structure of academic conferences.
Here, we introduce one such concept, hack weeks, and detail the underlying philosophical ideas along with experiences from events held in three different fields.

@@

Hack weeks are still a very young concept, and estimating the long-term impact of these events within the scientific communities they serve will require follow-up over multiple years to asses their effect on collaboration networks, career outcomes and adoption of new methods.
We have shown, however, that hack weeks provide an easy-to-implement, fairly low-cost method to introduce new technologies and methods into scientific fields on much shorter time scales than traditional teaching efforts can.
While we focus here on hack weeks in scientific fields, the concept could be extended to other areas, and is more generally useful in any area (1) where useful tools can be learned in short tutorials, (2) where results and outcomes can be produced on the timescale of a few days, and (3) that would benefit from collaborative approaches that cross traditional boundaries. Such areas could include for example the social sciences, music and art.

>>>>

The fast-paced changes in the computational and methodological landscape require that traditional fields of science rapidly adapt to new data analysis challenges.
To address these challenges, new types of workshops, including unconferences, hackathons and bootcamps have been developed in recent years in various scientific disciplines and now exist alongside with, and support the existing structure of academic conferences, formal classes, and other learning opportunities.
Here, we introduce one such concept, hack weeks, and detail the underlying philosophical ideas along with experiences from events held in three different fields.

@@

Hack weeks are still a very young concept, and estimating the long-term impact of these events within the scientific communities they serve will require follow-up over multiple years, to assess their effect on collaboration networks, career outcomes and adoption of new methods.
We have shown, however, that hack weeks provide an easy-to-implement, fairly low-cost method to introduce new technologies and methods into scientific fields on much shorter time scales than traditional teaching efforts.
While we focus here on hack weeks in scientific fields, the concept could be extended to other areas, and is more generally useful in any area (1) where useful tools can be learned in short tutorials, (2) where results and outcomes can be produced on the timescale of a few days, and (3) that would benefit from collaborative approaches that cross traditional boundaries. Such areas could include the social sciences, the humanities, as well as music and art.

### 2018-06-04 participants.tex

Traditional selection processes that rely heavily on internal heuristics of reviewers, especially those that consider characteristics peripheral to the evaluation criteria, are often fraught personal and structural biases~\cite[e.g.][]{sunstein2015wiser}. 
In order to maximize diversity and minimize bias, we advocate for a selection process that is as quantitative and transparent as possible~\cite{sunstein2015wiser}, enabling participants to hold organizers accountable for their selection decisions.
This requires laying out a definition of successful participation, what criteria must be met in order to maximize the likelihood of success, and how those criteria will be assessed given the information about the candidates selected during the application stage.

@@

If merit-based selection is part of the evaluation process, organizers face the crucial decision whether to assess merit blindly or not. Because human decision makers tend to be swayed by unrelated characteristics including name~\cite{bertrand2004} or gender~\cite{mossracusin2012}, an initial merit selection blinded to demographic characteristics can be an effective way to counteract certain biases related to internal heuristics. A merit selection could then be performed via scores given independently by members of the organizing committee based on a set of pre-defined, explicit selection criteria. This type of blinded procedure tends to reduce biases when committees would otherwise not consider diversity during their selection~\cite{bohnet2016}.

However, a blind selection purely on a figure of merit will be counterproductive if it excludes participants who might have had less exposure to certain technologies or fewer opportunities to learn certain skills: for example, requiring a minimum level of programming experience will likely disadvantage candidates who have had fewer opportunities to learn programming due to structural inequalities. 
Additionally, blinding has been found to have negative effects on diversity for committees that have already had a strong commitment to diversity, because they often correct for structural inequalities by considering demographic variables during merit selection~\cite{behaghel2015unintended}.
In this case, it may be beneficial to construct selection criteria that explicitly consider diversity and inclusivity (as Neuro Hack Week has done; see also SM, Section 4.1.5). 
Because systemic biases likely also enter at the application stage (where underrepresented groups may be less likely to apply) organizers should consider oversampling traditionally disenfranchised groups compared to the population of applicants.

No matter the selection procedures used, we encourage organizers to critically examine their cohort selection, experiment with new approaches, and routinely evaluate their procedure. For example, comparing demographic characteristics of the selected versus non-selected groups can unveil unintended biases during the merit-selection base and thus allow adjustments in the procedure to mitigate or fully remove these effects.

>>>>

Traditional selection processes that rely heavily on internal heuristics of reviewers, especially those that consider characteristics peripheral to the evaluation criteria, are often fraught with personal and structural biases~\cite[e.g.][]{sunstein2015wiser}.
In order to maximize diversity and minimize bias, we advocate for a selection process that is as quantitative and transparent as possible~\cite{sunstein2015wiser}, enabling participants to hold organizers accountable for their selection decisions.
This requires laying out a definition of successful participation, defining what criteria must be met in order to maximize the likelihood of success, and defining how those criteria will be assessed given the information about the candidates selected during the application stage.

@@

If merit-based selection is part of the evaluation process, organizers face the crucial decision whether to assess merit blinded to other applicant characteristics. Because human decision makers tend to be swayed by unrelated characteristics including name~\cite{bertrand2004} or gender~\cite{mossracusin2012}, an initial merit selection blinded to demographic characteristics can be an effective way to counteract certain biases. A merit selection could then be performed via scores given independently by members of the organizing committee based on a set of pre-defined, explicit selection criteria. This type of blinded procedure tends to reduce biases when committees would otherwise not consider diversity during their selection~\cite{bohnet2016}.

However, a blind selection based purely on an assessment of merit will be counterproductive if it excludes participants who might have had less exposure to certain technologies or fewer opportunities to learn certain skills: for example, requiring a minimum level of programming experience will likely disadvantage candidates who have had fewer opportunities to learn programming due to structural inequalities.
Additionally, blinding has been found to have negative effects on diversity for committees that  already have a strong commitment to diversity, because these circumstances allow to correct for structural inequalities by considering demographic variables during merit selection~\cite{behaghel2015unintended}.
In this case, it may be beneficial to construct selection criteria that explicitly consider diversity and inclusivity (as Neuro Hack Week has done; see also SM, Section 4.1.5).
Because systemic biases likely also enter at the application stage (where underrepresented groups may be less likely to apply) organizers should consider oversampling traditionally disenfranchised groups compared to the population of applicants.

No matter the selection procedures used, we encourage organizers to critically examine their cohort selection, experiment with new approaches, and routinely evaluate their procedure. For example, comparing demographic characteristics of the selected versus non-selected groups can unveil unintended biases during the merit-selection phase and thus allows adjustments in the procedure to mitigate or fully remove these effects.

### 2018-06-04 results.tex

One important question is whether participants from underrepresented groups thrive at hack weeks, or whether their full participation is impeded. Significant differences between minorities and non-minorities, even on a self-reported scale, for questions related to learning outcomes, teaching or network building would indicate that improvements in workshop facilitation and structure may be required to allow participants of all stripes to participate fully.
Based on our surveys, we find no significant dependence of self-reported learning outcomes on gender identity or race/ethnicity ($p > 0.0007$). Similarly, minority participants with respect to gender identity and race or ethnicity do not significantly differ in their answers with respect to teaching outcomes for any of the hack weeks ($p > 0.0007$). In neither case seem the attitudes toward building valuable connections at all three hack weeks or the value of their contributions to their hack teams differ with respect to gender or racial/ethnic identity. For GHW, there is a weak indication that participants from racial/ethnic minorities respond more positively when asked about building connections ($p=0.04$) with a medium to large effect size ($\phi_c = 0.4$; $\mathrm{dof}_{\phi_c} = 0.97$), while for AHW, responses regarding the value of contributions to hack teams may be spread more widely for participants from racial/ethnic minorities than for Caucasian participants ($p = 0.01$; $\phi_c = 0.4$, $\mathrm{dof}_{\phi_c} = 0.98$). These initial results indicate that future work on how demographics interact with hack week attendance may be fruitful.

We find that the hack weeks have been largely successful in promoting positive attitudes towards reproducibility and open science: at all three events, the majority reports that the hack week has made them more comfortable with open science  (GHW: 97\%; NHW: 95\%; AHW: 72\%, Figure \ref{fig:survey}, (h)), and more than 85\% of all participants (AHW: 86\%, GHW: 94\%, NHW: 95\%; Figure \ref{fig:survey}, (g)) put code or data created at the hack week into a public repository.
%The overall behaviour hints at how general conventions and attitudes differ in different fields.
%Astronomy shows the largest degree of openness toward open science, whereas our results indicate that open science is still fairly uncommon in the geosciences, with neuroscience falling in between.
%Similar attitudes are reflected when asking whether the hack week has made participants more comfortable with open science (Figure \ref{fig:survey}, (j)): again, geoscience shows the large improvement with over 97\% agreeing with this statement to some degree, followed by neuroscience (95\%) and astronomy (72\%).
%Overall, our results indicate that hack weeks are effective at addressing persisting doubts about making research open and reproducible. 
While the focus on open science is not necessarily a required component of a hack week, it aligns naturally with many of the goals and values commonly promoted at hack weeks like open-source software and data sharing. In some fields, especially where ethical issues around data sharing and privacy are relevant, this might not be a desired focus of the hack week, or might be replaced or augmented by a discussion of ethical considerations.

We caution the reader that our survey results should be read only as an initial indication of the hypotheses we proposed about the use and outcomes of hack weeks, in line with the surveys' exploratory nature. Our number of respondents is small and the effects likely subtle, thus the lack of significant differences may be caused by the small statistical power of our sample. The most important independent variable--attendance at the hack week--is something that our current survey design cannot account for. Moreover, self-reported learning outcomes are not an objective measure, because they are likely subject to internal biases of respondents. Future work will include a more refined survey design as well as attendance versus non-attendance as a control variable.

@@

\subsection*{Examples of Hack Week Outcomes}
\label{sec:outcomes}
\subsubsection*{Example 1: Astro Hack Week}
In 2015, a small team used AHW to found a new software project called Stingray\footnote{https://github.com/StingraySoftware/stingray} with the goal of providing implementations of time series analysis algorithms often used in astronomy.
Astro Hack Week enabled participants to seed a new collaboration around a software project needed by the larger community, facilitated by the collaborative environment at AHW.
Stingray has since matured into an enduring collaboration within the community with five active maintainers and four Google Summer of Code projects.
\subsubsection*{Example 2: Geo Hack Week}
In 2016, a GHW project team used Google Earth Engine to explore spatial patterns in climate, topography and population data with the goal of mapping the most suitable locations for renewable energy sites in the United States.
The team used machine learning algorithms in conjunction with the powerful hardware resources provided by Google Earth Engine\footnote{\url{http://georgerichardson.net/2017/04/10/searching-for-energy-in-a-random-forest/}}.
%George Richardson, one of the project leads, now works for a renewable resource company in Seattle.
\subsubsection*{Example 3: Neuro Hack Week}
Motion of study participants inside of the MRI machine is a major concern in neuroimaging studies.
During NHW 2016 one of the teams focused on a large and openly available data-set of MRI data from children\footnote{ABIDE: \url{http://preprocessed-connectomes-project.org/abide}}.
To test the effect of motion on the results, the team conducted an analysis in which both the number of experimental subjects included, as well as motion cut-off were varied.
%They tested both the split-half reliability of an analysis of brain connectivity, as well as an analysis that used machine learning to distinguish between brains of children with and without autism spectrum disorder.
The team (composed of four researchers from four institutions) continued to work on this project remotely after the end of NHW, and eventually published a paper describing these results in the open access journal Research Ideas and Outcomes \cite{leonard2017}.

>>>>

One important question is whether participants from underrepresented groups thrive at hack weeks, or whether their full participation is impeded. Significant differences between minorities and non-minorities, even on a self-reported scale, for questions related to learning outcomes, teaching or network building, would indicate that improvements in workshop facilitation and structure may be required to allow a wide range of participants to participate fully.
Based on our surveys, we find no significant dependence of self-reported learning outcomes on gender identity or race/ethnicity ($p > 0.0007$). Similarly, minority participants with respect to gender identity and race or ethnicity do not significantly differ in their answers with respect to teaching outcomes for any of the hack weeks ($p > 0.0007$). Attitudes towards building valuable connections at all three hack weeks or the value of their contributions to their hack teams also do not differ with respect to gender or racial/ethnic identity. For GHW, there is a weak indication that participants from racial/ethnic minorities respond more positively when asked about building connections ($p=0.04$) with a medium to large effect size ($\phi_c = 0.4$; $\mathrm{dof}_{\phi_c} = 0.97$), while for AHW, responses regarding the value of contributions to hack teams may be spread more widely for participants from racial/ethnic minorities than for Caucasian participants ($p = 0.01$; $\phi_c = 0.4$, $\mathrm{dof}_{\phi_c} = 0.98$). These initial results indicate that future work on how demographics interact with hack week attendance may be fruitful.

We find that the hack weeks have been largely successful in promoting positive attitudes towards reproducibility and open science: at all three events, the majority reports that the hack week has made them more comfortable with open science  (GHW: 97\%; NHW: 95\%; AHW: 72\%, Figure \ref{fig:survey}, (h)), and more than 85\% of all participants (AHW: 86\%, GHW: 94\%, NHW: 95\%; Figure \ref{fig:survey}, (g)) put code or data created at the hack week into a public repository.
%The overall behaviour hints at how general conventions and attitudes differ in different fields.
%Astronomy shows the largest degree of openness toward open science, whereas our results indicate that open science is still fairly uncommon in the geosciences, with neuroscience falling in between.
%Similar attitudes are reflected when asking whether the hack week has made participants more comfortable with open science (Figure \ref{fig:survey}, (j)): again, geoscience shows the large improvement with over 97\% agreeing with this statement to some degree, followed by neuroscience (95\%) and astronomy (72\%).
%Overall, our results indicate that hack weeks are effective at addressing persisting doubts about making research open and reproducible.
While the focus on open science is not necessarily a required component of a hack week, it aligns naturally with many of the goals and values commonly promoted at hack weeks, such as production of open-source software and data sharing. In some fields, especially where ethical issues around data sharing and privacy are relevant, this might not be a desired focus of the hack week, or might be replaced or augmented by a discussion of ethical considerations.

We caution the reader that our survey results should be read only as an initial indication of the hypotheses we proposed about the use and outcomes of hack weeks, in line with the surveys' exploratory nature. The number of respondents is small and the effects likely subtle. This means that the lack of significant differences may be caused by the small statistical power of our sample. The most important independent variable--attendance at the hack week--is something that our current survey design cannot account for. Moreover, self-reported learning outcomes are not an objective measure, because they are likely subject to internal biases of respondents. Future work will include a more refined survey design as well as attendance versus non-attendance as a control variable.

@@

\subsection*{Examples of Hack Week Outcomes}
\label{sec:outcomes}
\subsubsection*{Example 1: Astro Hack Week}
In 2015, a small team used AHW to found a new software project called Stingray\footnote{https://github.com/StingraySoftware/stingray} with the goal of providing implementations of time series analysis algorithms often used in astronomy.
Astro Hack Week enabled participants to seed a new collaboration around a software project needed by the larger community, facilitated by the collaborative environment at AHW.
Stingray has since matured into an enduring collaboration within the community with five active maintainers and four Google Summer of Code projects.
\subsubsection*{Example 2: Geo Hack Week}
In 2016, a GHW project team used Google Earth Engine to explore spatial patterns in climate, topography and population data with the goal of mapping the most suitable locations for renewable energy sites in the United States.
The team used machine learning algorithms in conjunction with the powerful hardware resources provided by Google Earth Engine\footnote{\url{http://georgerichardson.net/2017/04/10/searching-for-energy-in-a-random-forest/}}.
%George Richardson, one of the project leads, now works for a renewable resource company in Seattle.
\subsubsection*{Example 3: Neuro Hack Week}
Motion of study participants inside the MRI machine is a major concern in neuroimaging studies.
During NHW 2016, one of the teams focused on a large and openly available data-set of MRI data from children\footnote{ABIDE: \url{http://preprocessed-connectomes-project.org/abide}}.
To test the effect of motion on the results, the team conducted an analysis in which both the number of experimental subjects included, as well as motion cut-off, were varied.
%They tested both the split-half reliability of an analysis of brain connectivity, as well as an analysis that used machine learning to distinguish between brains of children with and without autism spectrum disorder.
The team (composed of four researchers from four different institutions) continued to work on this project remotely after the end of NHW, and eventually published a paper describing these results in the open access journal Research Ideas and Outcomes \cite{leonard2017}.
